Simulating High-Dimensional Multivariate Data

### 2020-10-26 010-background.Rmd
draft background

# Background and notation {#background}

@@

Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.

@@

@@

The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables $X$ and $Y$ and is given by

\begin{equation}
(\#eq:pearson)
\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
\end{equation}

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation adequately describes the dependency between the components. $\rho$'s utility in non-normal or non-linear associations is lacking (@MK01). Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted $\rho_s$) and Kendall's $\tau$. Define

\begin{equation}
(\#eq:spearman)
\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
\end{equation}

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. For continuous marginals, the measure works well due to zero probability of ties.

To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:

\begin{equation}
(\#eq:convertKendall)
r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
\end{equation}

@@

\begin{equation}
(\#eq:convertSpearman)
\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
\end{equation}

*Marginal-dependent bivariate correlation bounds*. 
Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
Yet not all multivariate simulation approaches obtain these bounds [for example REF].
Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].

The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, $[-1,-1]$. The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by [@BF17]:

\begin{equation}
(\#eq:frechet)
\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

@@

For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

\begin{equation}
(\#eq:spearmanRescaled)
\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
\end{equation}

@@

The crux of the method is construction of a Gaussian copula. This idea is well known [refs]. We chose a Gaussian copula among the many available copula functions primarily for two reasons: 1) the computationally convient closed expression to convert among the most common dependency measures when the margins are Gaussian and 2) the availability of high-performance multivariate normal simulators in R ('mvnfast' ref).

A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals. Moreover, for any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

@@

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$. The Gaussian copula is the one that corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$. (Since the marginals are standard normal, this ${\bf R}$ is also the correlation matrix). If $F_{{\bf R}}$ is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

@@

Sklar's Theorem (ref) guarantees that given inverse CDFs (quantile) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions. 
Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$. 
When an $Y_i$ is discrete, some care must be taken to define $F_i$. Letting

\begin{equation}
F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}
(\#eq:inverseCDF)
\end{equation}

\noindent ensures that $Y_i \sim F_i$.

Simulation of a multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula

>>>>

# Background {#background}

@@

@@

Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This results in `r choose(d,2)` pairwise dependencies to characterize and simulate.
Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.

@@

A central issue when describing multivariate distributinos is the choose of quantatitive description.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$ and is given by

\begin{equation}
(\#eq:pearson)
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
\end{equation}

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector the Pearson correlation completely describes the dependency between the components. 
For non-normal marginals and monotone (possibly non-linear) correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
Alternatively, in these settings, analysts often prefer rank-based correlatino measures to describe the degree of monotonic association.

Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$.

Formally, define

\begin{equation}
(\#eq:spearman)
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
\end{equation}

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
Spearman's $\rho$ has an appealing correspondence as the Pearson's correlaiton coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.

Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
By concordance we mean that orderings have the same direction if $X_i < X_j$, then $Y_i < Y_j$, and is determinded by the ranks of the values, not the values themselves (as in $\rho_P$).

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to our scheme to satistify properties SP1 and SP2, above.

This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficent, as the marginal distributions $F_i$ are fundamental to their caclutain.
But for bivariate normal vectors, however, the correspondence is well-known:

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

@@

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

These facts are also critical in our simulation algorithm to broaden the dependency measures supported by `bigsimr`.

*Discrete marginal considerations*.
Spearman's correlation for discrete marginal suffers some issues due to the nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ could be less than 1; @MB13).
One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
For two random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

\begin{equation}
(\#eq:spearmanRescaled)
\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}.
\end{equation}

Noting that with continuous marginals the rescaling returns $\rho_S$.
For discrete marginals with large or infinite support, computing adjustment factors $\sum_x p(x)^3$, $\sum_y p(y)^3$ over all large number of pairs becomes expensive (often violating desired property *SP1*).
And further the infinite sums must be approximately for count-valued data and potentially violating desired property *SP2*.

Also we'll note that for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
Let

\begin{equation}
F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}.
(\#eq:inverseCDF)
\end{equation}

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, the $\rho_P$ is not free to vary the entire range of possible correlations, $[-1,1]$.
These so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
This situatino gives strict restraints on the possible bounds and cannot be overcome through algorithm design.
In general the bounds are given by

\begin{equation}
(\#eq:frechet)
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

@@

@@

There is a strong connection of our simulatino strategy to Gaussian *copulas* (for a technical introduction see @Nelsen2007).
A copula is a distribution function on $[0,1]^d$ describing a random vector with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure and the study of copulas is important in statistical theory and practice.

For  any random vector ${\bf X}=(X_1, \ldots, X_d)$ with cumulative distribution function (CDF) $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

@@

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R}$.
But since the marginals are standardized to have unit variance, this ${\bf R}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

@@

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) *any* random vector can be obtained via transformations involving copula functions. 
For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .
